Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2018-09-13 article.Rnw
Make minor corrections

Parameter estimation requires the numerical maximization of
\autoref{eqn:ll-rcmp}. Since the derivatives of $\ell$ cannot be
obtained in closed forms, we use the \texttt{BFGS}
algorithm~\citep{Nocedal1995} as implemented in the function
\texttt{optim()} in \texttt{R} \citep{Rcore2017}. Standard errors for
the regression coefficients are obtained based on the observed
information matrix $\mathcal{I}(\bm{\theta})$, where
$\mathcal{I}(\bm{\theta}) = -\mathcal{H}(\bm{\theta})$ (Hessian matrix)
is computed numerically by central finite differences. Standard errors
for $\hat{\mu}_i$ and hence confidence intervals are obtained by using
the delta method~\citep[p. 89]{Pawitan2001}.

@@

To check the orthogonality property we compute the covariance matrix
between maximum likelihood estimators
$\hat{\bm{\theta}} = (\hat{\bm{\beta}}, \phi)$, obtained from the
observed information matrix,
Cov$(\hat{\bm{\theta}}) = \mathcal{I}^{-1}(\bm{\theta})$.
\autoref{fig:ortho-plot} shows the correlation between the regression
and dispersion parameter estimators for each simulation scenario, on the
correlation scale. The correlations are gnerally close to zero in all
cases suggesting the orthogonality property for the reparametrized
model. Interestingly, results in the first panel show that
corr($\hat{\beta}_{22}, \hat{\phi}$) is not very close to zero (values
between $-0.4$ and $0.2$) for strong overdispersion ($\phi = -1.6$).

@@

To illustrate the orthogonality, \autoref{fig:ortho-surf} displays
contour plots of the deviance surfaces for four simulated data sets of
size 1000 with $\mu=5$ and different values of the dispersion
parameters. The shapes of the deviance function show that the proposed
parametrization is better for both computation and asymptotic
(normal-based) inference. Furthermore, it is interesting to note that
the deviance function shape under strong overdispersion ($\phi=-1.6$) is
not as well behaved as the others; this is due to the inaccuracy of the
mean approximation \autoref{fig:approx-plot}(a). Although the
reparametrized model is always valid, $\mu$ and $\phi$ are orthogonal
only when $\mu$ is the expectation of the distribution, in other words,
when the approximation $\lambda - (\nu-1)/(2\nu)$ is accurate. This also
explains the results of corr$(\hat{\beta}_{22}, \phi)$ in the first
panel of \autoref{fig:ortho-plot}, since $\beta_{22}$ is negative and
associated with low counts.

@@

Following~\citet{Zeviani2014}, the linear predictor is given by
$$\eta_i = \beta_0 + \beta_{1j} \texttt{def}_i + \beta_{2j}
\texttt{def}_i^2$$
where $\eta_i = \log(\lambda_i)$ for the original COM-Poisson and
$\eta_i = \log(\mu_{ij})$ for COM-Poisson$_\mu$. The parameter
$\mu_{ij}$ is the (approximated) expected number of cotton bolls for the
$i$-th defoliation level ($i=$ 1: 0\%, 2: 25\%, 3: 50\%, 4: 75\% e 5:
100\%) and $j$-th growth stage ($j$ = 1: vegetative, 2: flower bud, 3:
blossom, 4: boll, 5: boll open), i.e. we have a different second order
effect of defoliation at each growth stage. The parameter estimates and
goodness-of-fit measures for the Poisson, COM-Poisson,
COM-Poisson$_\mu$, and quasi-Poisson regression models are presented in
\autoref{tab:coef-cotton}.

@@

For the analysis of this data set, based on the descriptive analysis
(\autoref{fig:desc-soy}), we proposed the following linear predictor
$$
\eta_i = \beta_0 + \gamma_i + \tau_j +
  \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
  \beta_{3j}\texttt{K}_k
$$
where $\gamma_i$ is the effect of $i$-th block ($i=$1: block II, 2:
block III, 3: block IV and 4: block V), $\tau_j$ is the effect of $j$-th
moisture level ($j=$1: 50\% and 2: 62.5\%), $\beta_{1}$ and $\beta_{2}$
give the baseline quadratic response over potassium levels \texttt{K}
($k=$1: 0.0, 2: 0.3, 3: 0.6, 4: 1.2, 5: 1.8 100mg dm$^{-3}$) and
$\beta_{3j}$ is interaction of the first order potassium effect
(\texttt{K}) for the $j$-th moisture level (\texttt{umid}); The
predictor $\eta_i$ is $\log(\lambda_i)$ for the original COM-Poisson and
$\log(\mu_{ij})$ for COM-Poisson$_\mu$. \autoref{tab:coef-soy} presents
the estimates, ratio between estimate and standard error, and
goodness-of-fit measures for the alternative models.

@@

For this data set we consider three models with linear predictors,
\begin{center}
\begin{minipage}{12cm}
Linear: $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i$\\
Quadratic: $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Cubic: $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}
where $\eta_i = \log(\lambda_i)$ for the original COM-Poisson and
$\eta_i = \log(\mu_{ij})$ for COM-Poisson$_\mu$.

\begin{table}[!ht]
\centering \small
\caption{Model fit measures and comparisons between predictors and
  models fitted to the nitrofen data.}
\label{tab:anova-ovos}
\begin{tabularx}{\textwidth}{lCCCCCrC}
  \toprule
 Poisson & np & $\ell$ & AIC & 2(diff $\ell$) & diff np & P($>\rchi^2$) & \\
 \midrule
<<anova-ovos1, results="asis">>=
@
\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  COM-Poisson & np & $\ell$ & AIC & 2(diff $\ell$) & diff np &
  P($>\rchi^2$) & $\hat{\phi}$ \\
  \midrule
<<anova-ovos2, results="asis">>=
@
\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  COM-Poisson$_\mu$ & np & $\ell$ & AIC & 2(diff $\ell$) & diff np &
  P($>\rchi^2$) & $\hat{\phi}$ \\
  \midrule
<<anova-ovos3, results="asis">>=
@
\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  Quasi-Poisson & np & QDev & AIC & F & diff np & P($>F$) & $\hat{\sigma}$ \\
  \midrule
<<anova-ovos4, results="asis">>=
@
 \bottomrule
\end{tabularx}
%\vspace{-1mm}

>>>>

Parameter estimation requires the numerical maximization of
\autoref{eqn:ll-rcmp}. Since the derivatives of $\ell$ cannot be
obtained in closed forms, we use the \texttt{BFGS}
algorithm~\citep{Nocedal1995} as implemented in the function
\texttt{optim()} in \texttt{R} \citep{Rcore2017}. Standard errors for
the regression coefficients are obtained based on the observed
information matrix $\mathcal{I}(\bm{\theta})$, where
$\mathcal{I}(\bm{\theta}) = -\mathcal{H}(\bm{\theta})$ (Hessian matrix)
is computed numerically by central finite differences. Standard errors
for $\hat{\eta}_i = \log(\hat{\mu}_i)$ and hence confidence intervals
are obtained by using the delta method~\citep[p. 89]{Pawitan2001}.

@@

To check the orthogonality property we compute the covariance matrix
between maximum likelihood estimators
$\hat{\bm{\theta}} = (\hat{\bm{\beta}}, \phi)$, obtained from the
observed information matrix,
Cov$(\hat{\bm{\theta}}) = \mathcal{I}^{-1}(\bm{\theta})$.
\autoref{fig:ortho-plot} shows the correlation between the regression
and dispersion parameter estimators for each simulation scenario, on the
correlation scale. The correlations are gnerally close to zero in all
cases suggesting the orthogonality property for the reparametrized
model. Interestingly, results in the first panel show that
Corr($\hat{\beta}_{22}, \hat{\phi}$) is not very close to zero (values
between $-0.4$ and $0.2$) for strong overdispersion ($\phi = -1.6$).

@@

To illustrate the orthogonality, \autoref{fig:ortho-surf} displays
contour plots of the deviance surfaces for four simulated data sets of
size 1000 with $\mu=5$ and different values of the dispersion
parameters. The shapes of the deviance function show that the proposed
parametrization is better for both computation and asymptotic
(normal-based) inference. Furthermore, it is interesting to note that
the deviance function shape under strong overdispersion ($\phi=-1.6$) is
not as well behaved as the others; this is due to the inaccuracy of the
mean approximation \autoref{fig:approx-plot}(a). Although the
reparametrized model is always valid, $\mu$ and $\phi$ are orthogonal
only when $\mu$ is the expectation of the distribution, in other words,
when the approximation $\lambda - (\nu-1)/(2\nu)$ is accurate. This also
explains the results of Corr$(\hat{\beta}_{22}, \phi)$ in the first
panel of \autoref{fig:ortho-plot}, since $\beta_{22}$ is negative and
associated with low counts.

@@

Following~\citet{Zeviani2014}, the linear predictor is given by
$$\eta_{ij} = \beta_0 + \beta_{1j} \texttt{def}_i + \beta_{2j}
\texttt{def}_i^2$$
where $\eta_{ij} = \log(\lambda_{ij})$ for the original COM-Poisson and
$\eta_{ij} = \log(\mu_{ij})$ for COM-Poisson$_\mu$. The parameter
$\mu_{ij}$ is the (approximated) expected number of cotton bolls for the
$i$-th defoliation level ($i=$ 1: 0\%, 2: 25\%, 3: 50\%, 4: 75\% e 5:
100\%) and $j$-th growth stage ($j$ = 1: vegetative, 2: flower bud, 3:
blossom, 4: boll, 5: boll open), i.e. we have a different second order
effect of defoliation at each growth stage. The parameter estimates and
goodness-of-fit measures for the Poisson, COM-Poisson,
COM-Poisson$_\mu$, and quasi-Poisson regression models are presented in
\autoref{tab:coef-cotton}.

@@

For the analysis of this data set, based on the descriptive analysis
(\autoref{fig:desc-soy}), we proposed the following linear predictor
$$
\eta_{ijk} = \beta_0 + \gamma_i + \tau_j +
  \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
  \beta_{3j}\texttt{K}_k
$$
where $\gamma_i$ is the effect of $i$-th block ($i=$1: block II, 2:
block III, 3: block IV and 4: block V), $\tau_j$ is the effect of $j$-th
moisture level ($j=$1: 50\% and 2: 62.5\%), $\beta_{1}$ and $\beta_{2}$
give the baseline quadratic response over potassium levels \texttt{K}
($k=$1: 0.0, 2: 0.3, 3: 0.6, 4: 1.2, 5: 1.8 100mg dm$^{-3}$) and
$\beta_{3j}$ is interaction of the first order potassium effect
(\texttt{K}) for the $j$-th moisture level (\texttt{umid}); The
predictor $\eta_{ijk}$ is $\log(\lambda_{ijk})$ for the original
COM-Poisson and $\log(\mu_{ijk})$ for
COM-Poisson$_\mu$. \autoref{tab:coef-soy} presents the estimates, ratio
between estimate and standard error, and goodness-of-fit measures for
the alternative models.

@@

For this data set we consider three models with linear
predictors,\\[\parskip]
\hspace*{1.5\parskip}
\begin{tabular}{ll}
  Linear:     & $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i$,\\
  Quadratic: & $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i +
               \beta_2 \texttt{dose}_i^2$ e\\
  Cubic:     & $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i +
               \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$,
\end{tabular}

where $\eta_i = \log(\lambda_i)$ for the original COM-Poisson and
$\eta_i = \log(\mu_i)$ for COM-Poisson$_\mu$.

\begin{table}[!ht]
\centering \small
\caption{Model fit measures and comparisons between predictors and
  models fitted to the nitrofen data.}
\label{tab:anova-ovos}
\begin{tabularx}{\textwidth}{lCCCCCrC}
  \toprule
 Poisson & $\#p$ & $\ell$ & AIC & 2(diff $\ell$) & diff $\#p$ & P($>\rchi^2$) & \\
 \midrule
<<anova-ovos1, results="asis">>=
@
\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  COM-Poisson & $\#p$ & $\ell$ & AIC & 2(diff $\ell$) & diff $\#p$ &
  P($>\rchi^2$) & $\hat{\phi}$ \\
  \midrule
<<anova-ovos2, results="asis">>=
@
\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  COM-Poisson$_\mu$ & $\#p$ & $\ell$ & AIC & 2(diff $\ell$) & diff $\#p$ &
  P($>\rchi^2$) & $\hat{\phi}$ \\
  \midrule
<<anova-ovos3, results="asis">>=
@
\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  Quasi-Poisson & $\#p$ & QDev & AIC & F & diff $\#p$ & P($>F$) & $\hat{\sigma}$ \\
  \midrule
<<anova-ovos4, results="asis">>=
@
 \bottomrule
\end{tabularx}
%\vspace{-1mm}
